\MakeUppercase{The Structure of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-07-04 k-planar.tex
Revised page 2

We now generalize \thmref{k-planar} to a more abstract setting.  For two sets $A$ and $B$, we use $A\ominus B := (A\cup B)\setminus(A\cap B)$ to denote the symmetric difference between $A$ and $B$.  We begin with a fairly standard lemma about normalized tree decompositions:

@@

Let $G$ be a graph and let $(\mathcal{L}:=\langle V_1,\ldots,V_h\rangle,\mathcal{P}:=(Q_x: x\in V(H)))$ be a layered partition of $G$ of layered width $\ell$ where the treewidth of the quotient graph $H=G/\mathcal{P}$ is at most $t$.  In this section we define a hierarchical structure on $G$ that can be obtained from $(\mathcal{L},\mathcal{P})$.

\begin{lem}\thmlabel{generalized-tripod}
  There exists a tree $T$ rooted at a node $x_0$ where each node $x\in V(T)$ is associated with a subset $Y_x\subseteq V(G)$ that defines several related sets $V_x = \bigcup_{y\in V(T_x)} Y_y$, $F_x=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x=V_x\cup F_x$ that satisfy the following properties:
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width (with respect to $\mathcal{L}$) at most $t\ell$.
    \item\ylabel{separator} There is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} $x$ has $t'\le t$ strict $T$-ancestors $a_1,\ldots,a_{t'}$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{lem}

Before proving \thmref{generalized-tripod} we point two properties that are immediately implied by it:

@@

Property~\yref{y-subsets} follows from the fact that $V_x$ is the union of several sets, one of which is $Y_x$.  Property~\yref{containment-i} follows from the definitions of $V_x$ and the fact that the subtree $V(T_x)\subseteq V(T_a)$. To show Property~\yref{containment-ii} first note that, by \yref{containment-i} it suffices to consider vertices in $w\in F_x=N_x\setminus V_x$. By definition, every vertex $w\in F_x$ is adjacent, in $G$, to a vertex $v\in V_x$.  By \yref{containment-i}, $v\in V_a$, so $w$ is either in $V_a$ or $w$ satisfies the condition $vw\in E(G)$, $v\in V_a$, and $w\not\in V_a$, so $w\in F_a$.  In either case $w\in N_a=V_a\cup F_a$.  Note that none of \yref{y-subsets}--\yref{containment-ii} depends on \yref{ancestor-edge} (which is important, since \yref{containment-i} is used to establish \yref{ancestor-edge} below).

\begin{proof}[Proof of \thmref{generalized-tripod}]
  Apply \lemref{nice-decomposition} to $I$ to obtain a tree-decomposition $(B_x: x\in V(T))$ of $I$ where $T$ is rooted at some node $x_0$ and that satisifies Properties~\tref{first}--\tref{last}. (We skip the introduction of the isomorphic graph $H'$ and assume from the outset that $V(I)\subseteq V(T))$.)

  For each $v\in V(G)$, let $q(v)$ denote the unique node $q\in V(I)$ such that $v\in Q_q$ and let $b(v)$ denote the root of the subtree $T[v]:=T[\{x\in V(T):q(v)\in B_x\}]$. For each $x\in V(T)$, define $Y_x=\{v\in V(G):c(v)=x\}$.  
  Clearly, $\mathcal{Y}:=(Y_x:x\in V(T))$ is a partition of $V(G)$.  Furthermore, for any $i\in\{1,\ldots,h\}$ 
  \[  |Y_x\cap V_i| = \sum_{y\in C_x}|Q_y\cap V_i|\le |C_x|\ell \le t\ell ,
  \] 
  so $\mathcal{Y}$ has layered width at most $t\ell$ with respect to $\mathcal{L}$.  This establishes \yref{thickness}.

@@

  To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}$ for some $a\in V(T)$, $a\neq x$.  Since $v\in Y_{x'}$ and $w\in Y_a$, $q(v)=x'$ and $q(w)=a$, so $x'\in Q_v$ and $w\in Q_a$.  Therefore $H$ contains the edge $ax'$.  By \tref{ancestor-edge}, $a$ is an ancestor of $x'$ or $x'$ is an ancestor of $a$.  Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $x'=z_0,\ldots,z_r=a$ be the path, in $T$ from $x'$ to $a$.  For each $i\in\{0,\ldots,r\}$, at least one of $a$ or $x'$ is in $C_{z_i}$.  However, $b(x')=x'$, so $x'$ is not contained in $C_{x_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in C_{x_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $C_x$.
  Property~\yref{ancestor-edge} now follows from the fact that $|C_x|\le t$.
\end{proof}

>>>>

We now generalize \thmref{k-planar} to a more abstract setting.  For two sets $A$ and $B$, we use $A\ominus B := (A\cup B)\setminus(A\cap B)$ to denote the symmetric difference between $A$ and $B$.  For a rooted tree $T$ and any node $x\in V(T)$, we define $T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}]$ to be the maximal subtree of $T$ rooted at $x$.
We begin with a fairly standard lemma about normalized tree decompositions:

@@

For the remainder of this section, $G$ is a graph and $(\mathcal{L}:=\langle V_1,\ldots,V_h\rangle,\mathcal{P}:=(Q_x: x\in V(H)))$ is a layered partition of $G$ of layered width $\ell$ where the treewidth of the quotient graph $H=G/\mathcal{P}$ is at most $t-1$.  The following lemma shows how to use $(\mathcal{L},\mathcal{P})$ to define a hierarchical decomposition of $G$.

\begin{lem}\thmlabel{generalized-tripod}
  There exists a tree $T$ rooted at a node $x_0$ where each node $x\in V(T)$ is associated with a subset $Y_x\subseteq V(G)$ that defines several related sets $V_x = \bigcup_{y\in V(T_x)} Y_y$, $F_x=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x=V_x\cup F_x$ that satisfy the following properties:
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width (with respect to $\mathcal{L}$) at most $\ell$.
    \item\ylabel{separator} There is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} $x$ has $t'\le t$ strict $T$-ancestors $a_1,\ldots,a_{t'}$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{lem}

Before proving \thmref{generalized-tripod} we point out more properties that are immediately implied by it:

@@

Property~\yref{y-subsets} follows from the fact that $V_x$ is the union of several sets, one of which is $Y_x$.  Property~\yref{containment-i} follows from the definition of $V_x$ and the fact that $V(T_x)\subseteq V(T_a)$. To show Property~\yref{containment-ii} first note that, by \yref{containment-i} it suffices to consider vertices in $w\in F_x=N_x\setminus V_x$. By definition, every vertex $w\in F_x$ is adjacent, in $G$, to a vertex $v\in V_x$.  By \yref{containment-i}, $v\in V_a$, so $w$ is either in $V_a$ or $w$ satisfies the condition $vw\in E(G)$, $v\in V_a$, and $w\not\in V_a$, so $w\in F_a$.  In either case $w\in N_a=V_a\cup F_a$.  Note that none of \yref{y-subsets}--\yref{containment-ii} depends on \yref{ancestor-edge} (which is important, since \yref{containment-i} is used to establish \yref{ancestor-edge} in the following proof).

\begin{proof}[Proof of \thmref{generalized-tripod}]
  Apply \lemref{nice-decomposition} to $H$ to obtain a tree-decomposition $(B_x: x\in V(T))$ of $H$ where $T$ is rooted at some node $x_0$ and that satisifies Properties~\tref{first}--\tref{last}. (We skip the introduction of the isomorphic graph $H'$ and assume from the outset that $V(H)\subseteq V(T))$.)

  For each $v\in V(G)$, let $q(v)$ denote the unique node $q\in V(H)$ such that $v\in Q_q$.  For each $x\in V(T)$, define
  \[
      Y_x = \begin{cases} Q_x & \text{if $x\in V(H)$} \\
                          \emptyset & \text{otherwise} .
            \end{cases}
  \]
   % 
   % 
   % and let $b(v)$ denote the root of the subtree $T[v]:=T[\{x\in V(T):q(v)\in B_x\}]$. For each $x\in V(T)$, define $Y_x=\{v\in V(G):c(v)=x\}$.  
  Clearly, $\mathcal{Y}:=(Y_x:x\in V(T))$ is a partition of $V(G)$ and, for any $i\in\{1,\ldots,h\}$ $|Y_x\cap V_i|\le \ell$.
  so $\mathcal{Y}$ has layered width at most $t\ell$ with respect to $\mathcal{L}$.  This establishes \yref{thickness}.

@@

  To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}=Q_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}=Q_a$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $x'=z_0,\ldots,z_r=a$ be the path, in $T$ from $x'$ to $a$.  For each $i\in\{0,\ldots,r\}$, at least one of $a$ or $x'$ is in $C_{z_i}$.  However, by \tref{subtree-root} $x'$ is not contained in $C_{x_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in C_{x_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $C_x$.
  Property~\yref{ancestor-edge} now follows from the fact that $|C_x|\le t+1$.
\end{proof}
